Model-agnostic Interpretable Machine Learning

### 2021-08-23 2021-bips-practice-phd-defence/2021-bips-practice-phd-defence.Rmd
finish slides

---
title: "Model-agnostic Interpretable Machine Learning"
author: "Christoph Molnar"
date: "2021-08"
institute: "BIPS bremen"
output:
  xaringan::moon_reader
---

@@

# Introduction

@@

Interpretable ML has 200 year old roots, but consolidation only started in 2016.

TODO: Web Of Science for keywords XAI and IML (written out)

## Why Interpretability?

```{r, out.width = 500}
knitr::include_graphics("../images/explain.png")
```

@@

@@

## Model-agnostic Interpretation

@@

@@

How does a feature influence the prediction for an individual observation **on average**?

@@

The theoretical construct:
$$ PD(x) = \mathbb{E}_{X_{C}}[\hat{f}(x, X_{C})] $$

@@

The algorithm (simples form): CITE Breiman

@@

## Difference between model-agnostic and model-specific interpetation

* Analyzing components vs. analyzing behavior
* Components not very powerful, unless they directly describe behavior
* For complex model, components do not directly tell us about behavior of model
* Behavior more comparable

@@

@@

## Summary of Thesis

@@

# Inference of Feature Effects and Importance in Machine Learning

## Goal of Paper

Treat permutation feature importance (PFI) and partial dependence plots (PDP) as statistical estimators.

@@

@@

TODO: Draw image which shows link between model and DGP

## Status-Quo

- For inference, practitioner rely on interpretable models
- But: situations where more flexible models match DGP better (measured in test loss)
- Interpretation possible with PDP and PFI (among others)
- However only describes the model behavior
- Question: Can we link PD and PFI to the DGP? and even do inference?

@@

- Hope: Lift PDP and PFI to status of statistical estimator
- Needed: estimand. A definition of what it is supposed to estimate in the DGP (not the model)
- Idea: PDP and PFI are expectation of a function
- Assumption: True function $f$ exists in DGP
- Then: We can apply same interpretation tools to DGP. Very simple.

TODO: simple image with DGP + model. External tools, PDP and PFI, applied externally and compared.

@@

@@

## Model-PD and Model-PFI

Show definition plus confidence intervals plus image

@@

Show definition plus confidence intervals plus image

But there is a problem when estimating this variance

## Variance correction

>>>>

---
title: "Model-agnostic Interpretable Machine Learning"
author: "Christoph Molnar"
date: "2021-08"
institute: "BIPS bremen"
output:
  xaringan::moon_reader:
    css: ["default", "default-fonts", "references.css"]
---

@@

@@

@@

---
class: center, middle
background-color: darkgrey

@@

@@

How does a feature influence the prediction for an **individual data point**?

@@

How does a feature influence the prediction **on average**?

@@

The theoretical construct:
$$PD(x) = \mathbb{E}_{X_{C}}[\hat{f}(x, X_{C})]$$

@@

The algorithm (simples form):

@@

---
class: middle, center
background-color: darkgrey

@@

---
background-color: lightgrey

@@

---
background-image:url(../images/thesis-summary.jpeg)
background-size: contain

@@

---
class: center, middle
background-color: darkgrey

# Treating PDP and PFI as Statistical Estimators

@@

```{r, out.width = 500}
knitr::include_graphics("../images/dgp-model-mapping.jpeg")
```

@@

@@

```{r, out.width = 500}
knitr::include_graphics("../images/no-dgp-model-mapping.jpeg")
```

???
- For inference, practitioner rely on interpretable models
- But: situations where more flexible models match DGP better (measured in test loss)
- Interpretation possible with PDP and PFI (among others)
- However only describes the model behavior
- Question: Can we link PD and PFI to the DGP? and even do inference?
- Hope: Lift PDP and PFI to status of statistical estimator
- Needed: estimand. A definition of what it is supposed to estimate in the DGP (not the model)
- Idea: PDP and PFI are expectation of a function
- Assumption: True function $f$ exists in DGP
- Then: We can apply same interpretation tools to DGP. Very simple.

## Linking External Tools to DGP

```{r, out.width = 500}
knitr::include_graphics("../images/external-dgp-model-mapping.jpeg")
```

@@

```{r, out.width = 900}
knitr::include_graphics("../images/pdp-pfi-deviations.png")
```

@@

```{r, out.width = 640}
knitr::include_graphics("../images/pdp-pfi-ci.png")
```
---

## One Problem: Variance Underestimation

```{r, out.width = 500}
knitr::include_graphics("../images/overlapping-training-data.jpeg")
```

-> Variance correction needed

@@

class: small

## References

- Molnar, C., Casalicchio, G., & Bischl, B. (2018). iml: An R package for Interpretable Machine Learning. Journal of Open Source Software, 3, 786.
- Molnar, C.. (2018). Interpretable Machine Learning - A Guide for Making Black Box Models Explainable.
- Molnar, C., Casalicchio, G., & Bischl, B. (2019). Quantifying Model Complexity via Functional Decomposition for Better Post-Hoc Interpretability. In Joint European Conference on Machine Learning and Knowledge Discovery in Databases (pp. 193–204).
- Dandl, S., Molnar, C., Binder, M., & Bischl, B. (2020). Multi-objective counterfactual explanations. In Bäck T. et al. (eds) Parallel Problem Solving from Nature – PPSN XVI. PPSN 2020. Lecture Notes in Computer Science (pp. 448–469).
- Molnar, C., König, G., Bischl, B., & Casalicchio, G. (2020). Model-agnostic Feature Importance and Effects with Dependent Features–A Conditional Subgroup Approach. arXiv preprint arXiv:2006.04628.
- Molnar, C., König, G., Herbinger, J., Freiesleben, T., Dandl, S., Scholbeck, C., Casalicchio, G., Grosse-Wentrup, M., & Bischl, B. (2020). Pitfalls to Avoid when Interpreting Machine Learning Models. In ICML workshop XXAI: Extending Explainable AI Beyond Deep Models and Classifiers.
- Molnar, C. (2013). Recursive partitioning by conditional inference. Department of Statistics, University of Munich: Munich, Germany.
- Molnar, C., Casalicchio, G., & Bischl, B. (2020). Interpretable Machine Learning - A Brief History, State-of-the-Art and Challenges. arXiv preprint arXiv:2010.09337.
- Molnar, C., Freiesleben, T., König, G., Casalicchio, G., Wright, M., & Bischl, B. (2021), Relating Partial Dependence Plot and Permutation Feature Importance to the Data-Generating Process (Working Title). Not even on arxiv yet.
- Casalicchio, G., Molnar, C., & Bischl, B. (2018). Visualizing the feature importance for black box models. In Joint European Conference on Machine Learning and Knowledge Discovery in Databases (pp. 655–670).
- Scholbeck, C., Molnar, C., Heumann, C., Bischl, B., & Casalicchio, G. (2019). Sampling, intervention, prediction, aggregation: A generalized framework for model-agnostic interpretations. arXiv preprint arXiv:1904.03959.
- König, G., Molnar, C., Bischl, B., & Grosse-Wentrup, M. (2021). Relative Feature Importance. In 2020 25th International Conference on Pattern Recognition (ICPR) (pp. 9318–9325).
